A boundary-integral framework to simulate \\ the viscous erosion of a porous medium

### 2018-04-06 docs/methods/erosion_methods.tex
work on intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Flow-induced erosion deteriorates and reshapes solid material over a range of scales found in nature, from large geological formations~\cite{han1969, Rothman2012, coh-dev-sey-yi-szy-rot2015}, down to human-scale features and patterns~\cite{daerr2003erosion}. Though less visible, these same forces are at work at the very smallest scales, slowly deteriorating the individual constituents of porous media (e.g.~solid, sand, or clay) or biological structures like plaque and biofilms~\cite{pic-van-hei2000, sha2002, gro-gij-van-fer-hat-van-yua-wen2007}. Motivated by such applications, we develop numerical methods to simulate fluid-mechanical erosion of multiple solid bodies in Stokes flow---the most relevant regime for groundwater and bio-fluid applications.

Recent work in the high-Reynolds number regime has highlighted the importance
of the feedback between flow and morphology that occurs during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, hewett2017evolution, moore2017riemann}, and a similar feedback occurs in the related processes of dissolution and melting~\cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, hewett2017pear, claudin2017dissolution}. In such cases, the flow alters the shape of the immersed structures, which in turn modifies the surrounding flow, and so on. Collectively, such processes have been referred to as `sculpting with flow'~\cite{ristroph2018sculpting}. From this viewpoint, the example of an eroding porous medium presents new opportunities and challenges. First, the medium itself is composed of many solid constituents, so multi-body interactions, especially close interactions, must be considered. Second, the long-range interactions facilitated by Stokes flow even more strongly couples the shape evolution of neighboring bodies, making their interactions even more important.

The methods developed here will allow us to simulate the erosion of many
bodies, as demonstrated in Fig.~\ref{fig:50bodies} for the case of 50
bodies (of randomly assigned initial sizes and positions). Here, the
erosion of solid material arises from the mechanical stress induced by
the surrounding flow. In particular, we take the local rate of erosion
to be proportional to the magnitude of the shear stress---a law that has
been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}.
As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size
of the bodies, but also alters their shapes considerably. Notice, in
particular, that many bodies develop corner-like features and nearly
flat faces, with straight channels intervening between them. In the
figure, color represents the vorticity of the surrounding flow, which
reduces to shear on solid boundaries and thus provides a convenient tool
for visualizing local erosion rates. To perform this simulation of
Stokesian erosion of a porous media, two fast and accurate methods are
employed---one to solve the fluid equations and the other to integrate
the eroding grains.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The fluid equations are reformulated in terms of a boundary integral
equation (BIE).  Since the boundary conditions are Dirichlet, we use a
double-layer potential formulation so that the resulting BIE is
second-kind.  This second-kind BIE is discretized with the spectrally
accurate trapezoid rule~\cite{tre-wei2014} which results in a
well-conditioned dense linear system for an unknown density function.
This linear system is solved using fast-multipole-accelerated
GMRES~\cite{saa-sch1986}.  Once the density function is computed, the
erosion rate (shear stress) is computed with a spectrally accurate
odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure
and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies})
with an upsampled trapezoid rule in the fluid bulk to assist with
visualizing the effect of erosion in porous media.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Flow-induced erosion deteriorates and reshapes solid material over a
range of scales found in nature, from large geological
formations~\cite{han1969, Rothman2012, coh-dev-sey-yi-szy-rot2015}, down
to human-scale features and patterns~\cite{daerr2003erosion}. Though
less visible, these same forces are at work at the very smallest scales,
slowly deteriorating the individual constituents of porous media
(e.g.~solid, sand, or clay) or biological structures like plaque and
biofilms~\cite{pic-van-hei2000, sha2002, gro-gij-van-fer-hat-van-yua-wen2007}. Motivated by such applications, we
develop numerical methods to simulate fluid-mechanical erosion of
multiple solid bodies in a Stokes flow---the most relevant regime for groundwater and bio-fluid applications.

Recent work in the high-Reynolds number regime has highlighted the
importance of the feedback between flow and morphology that occurs
during erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013,
hewett2017evolution, moore2017riemann}, and a similar feedback occurs in
the related processes of dissolution and melting~\cite{Huang2015,
kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion,
hewett2017pear, claudin2017dissolution}. In such cases, the flow alters
the shape of the immersed structures, which in turn modifies the
surrounding flow, and so on. Collectively, such processes have been
referred to as `sculpting with flow'~\cite{ristroph2018sculpting}. From
this viewpoint, the example of an eroding porous medium presents new
opportunities and challenges. First, the medium itself is composed of
many solid constituents, so multi-body interactions, especially close
interactions, must be considered. Second, the long-range interactions
facilitated by Stokes flow makes all the interactions between bodies important.

The methods developed here will allow us to simulate the erosion of many
bodies, as demonstrated in Fig.~\ref{fig:50bodies} for the case of 50
bodies (of randomly assigned initial sizes and positions). Here, the
erosion of solid material arises from the mechanical stress induced by
the surrounding flow. In particular, we take the local rate of erosion
to be proportional to the magnitude of the shear stress---a law that has
been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}.
As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size
of the bodies, but also alters their shapes considerably. Notice, in
particular, that many bodies develop corner-like features and nearly
flat faces, with straight channels intervening between them. In the
figure, color represents the vorticity of the surrounding flow, which
reduces to shear on solid boundaries and thus provides a convenient tool
for visualizing local erosion rates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To perform this simulation of Stokesian erosion of a porous media, two
fast and accurate methods are employed---one to solve the fluid
equations and the other to integrate the eroding grains.  The fluid
equations are reformulated in terms of a boundary integral equation
(BIE).  Since the boundary conditions are Dirichlet, we use a
double-layer potential formulation so that the resulting BIE is
second-kind.  This second-kind BIE is discretized with the spectrally
accurate trapezoid rule~\cite{tre-wei2014} which results in a
well-conditioned dense linear system for an unknown density function.
This linear system is solved using fast-multipole-accelerated
GMRES~\cite{saa-sch1986}.  Once the density function is computed, the
erosion rate (shear stress) is computed with a spectrally accurate
odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure
and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies})
with an upsampled trapezoid rule in the fluid bulk to assist with
visualizing the effect of erosion in porous media.
